---
title: "Week 2 Notes"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
***
## Lattice plotting system

Lattice plotting system is implemented using the following packages:

* *lattice*: contains code for producing Trellis graphics, which are independent of the *"base"* graphics system; includes function like **xplot, bwplot, levelplot**

* *grid*: implements a different graphing system independent of the *"base"* system; the *lattice* package builds on top of *grid*

  - We seldom call functions from the *grid* package directly
  
* The lattice plotting system does not have a "two-phase" aspect with separate plotting and annotations like in base plotting.

* All plotting/annotations is donoe at once with a single function call.

## Lattice Fuctions

### xyplot

```{r}
library(lattice)
library(datasets)
## Simple scatterplot

xyplot(Ozone ~ Wind, data = airquality)
```

>Convert Month toa factor variable

```{r}
library(lattice)
library(datasets)
## Simple scatterplot
airquality <- transform(airquality, Month=factor(Month))
xyplot(Ozone ~ Wind | Month, data = airquality, layout=c(5,1))

```

